---
title: "Identifying Unique Locations"
author: "Mazama Science"
date: "August 18, 2020"
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem

When working with environmental monitoring time series, one of the first things
you have to do is create unique identifiers for each individual time series. In 
an ideal world, each environmental time series would have both a 
`locationID` and a `sensorID` that uniquely identify the spatial location and 
specific instrument making measurements. A unique `timeseriesID` could
be produced as `locationID_sensorID`. Location metadata associated with each
time series would contain basic information needed for downstream analysis
including at least:

`timeseriesID, locationID, sensorID, longitude, latitude, ...`

* Multiple sensors placed at a location could be be grouped by `locationID`.
* An extended time series for a mobile sensor would group by `sensorID`.
* Maps would be created using `longitude, latitude`.
* Time series would be accessed from a secondary `data` table with `timeseriesID`.

Unfortunately, we are rarely supplied with a truly unique and truly spatial 
`locationID`. Instead we often use `sensorID` or an associated non-spatial
identifier as a standin for `locationID`.

Complications we have seen include:

* GPS-reported longitude and latitude can have _jitter_ in the fourth or fifth 
decimal place making it challenging to use them to create a unique `locationID`.
* Sensors are sometimes _repositioned_ in what the scientist considers the "same 
location".
* Data for a single sensor goes through different processing pipelines using
different identifiers and is later brought together as two separate timeseries.
* The radius of what constitutes a "single location" depends on the 
instrumentation and scientific question being asked.
* Deriving location-based metadata from spatial datasets is computationally 
intensive unless saved and identified with a unique `locationID`.
* Automated searches for spatial metadata occasionally produce incorrect results
because of the non-infinite resolution of spatial datasets and must be corrected
by hand.

## MazamaLocationUtils

The **MazamaLocationUtils** package provides a solution to these problems
by storing spatial metadata in simple tables in a standard directory. These 
tables will be referred to as _collections_. Location lookups can be performed 
with geodesic distance calculations where a location is assigned to a pre-existing
_known location_ if it is within `radius` meters. These will be extremely fast.

If no previously _known location_ is found, the relatively slow (seconds)
creation of a new _known location_ metadata record can be performed and then 
added to the growing collection.

For collections of stationary environmental monitors that only number in the 
thousands, this entire _collection_ (_i.e._ "database") can be stored as either a 
`.rda` or `.csv` file and will be under a megabyte in size making it fast to 
load. This small size also makes it possible to store multiple _known location_ 
files, each created with different locations and different radii to address 
the needs of different scientific studies.

While initially exploring spatial metadata for RAWS sites, the
`MazamaLocationUtils::table_findOverlappingLocations()` function will identify
locations that seem too close to be considered "unique". It will ultimately
be up to the user of the data to decide what to do with "overlapping" time
series.

The following examples demonstrate how to explore RAWS location metadata using
functionality from **MazamaLocationUtils**.

## Working with data from cefa.dri.edu/raws

Well formamtted archival RAWS data is avaialable from https://cefa.dri.edu/raws.

### Spatial metadata

Metadata containing station IDs and location information can be accessed using
the `fw13_createMetadtat()` function

```{r fw13_createMetadata, message = FALSE}
library(RAWSmet)

library(MazamaSpatialUtils)
setSpatialDataDir("~/Data/Spatial")

meta <- fw13_createMetadata(verbose = TRUE)
head(meta)
```

This dataframe of `r nrow(meta)` records contains unique identifiers and locations.

```{r are_locations_unique}
nrow(meta)
meta %>% dplyr::select(nwsID, longitude, latitude) %>% dplyr::n_distinct()
```

We can use the **MazamaLocationUtils** to discover those locations that
are too close to be considered unique "known locations".

If we assume that stations spaced <200 meters apart are measuring the same
parcel of air, we can say that unique "known locations" for RAWS stations should
have a radius of 100m. We can use the following code see if this is the case:

```{r tooClose_100}
tooCloseTbl <-
  MazamaLocationUtils::table_findOverlappingLocations(meta, radius = 100)

print(tooCloseTbl)

# Extract siteName from meta dataframe
for ( i in seq_len(nrow(tooCloseTbl)) ) {
  rows <- as.numeric(tooCloseTbl[i, 1:2])
  cat(sprintf("\n%5.1f meters apart:\n", tooCloseTbl$distance[i]))
  print(meta[rows, c('longitude', 'latitude', 'siteName')])
}
```

### Timeseries data 1 -- Overlapping timeseries

Given that many of these `nwsID` pairs share the same `siteName`, we can create
a timeseries plot do see which of two possible situations we all into:

1. two RAWS instruments with overlapping measurements that should be considered 
separate time series
2. two RAWS instruments with non-overlapping measurements that can be considered
a single time series

Obtain timeseries objects (with separate `meta` and `data` dataframes) using the
`nwsID` obtained from each record:

```{r timeseries_identical}
# CAMP CREEK stations are separated by 56.7 m, row #4 from tooCloseTbl
#    ...
#    4  1597  1598     56.7

nwsID_1 <- meta$nwsID[1597]
nwsID_2 <- meta$nwsID[1598]

# Get timeseries data
CampCreek_1 <- fw13_createTimeseriesObject(nwsID_1, meta)
CampCreek_2 <- fw13_createTimeseriesObject(nwsID_2, meta)

# Each timeseries object consists of 'meta' and 'data'.
# We are only interested in the 'data' part:

head(CampCreek_1$data)

# ----- Plot them separately -----

# Hint:  use pch = '.' to speed up graphical rendering

layout(matrix(seq(2)))
plot(CampCreek_1$data[,c('datetime', 'temperature')], 
     col = 'black', pch = '.', main = "Camp Creek 1")
plot(CampCreek_2$data[,c('datetime', 'temperature')], 
     col = 'orange', pch = '.', main = "Camp Creek 2")
layout(1)

# ----- Plot them together -----

# Set up x- and y-limits to cover both timeseries
xlim <- range(c(
  range(CampCreek_1$data$datetime, na.rm = TRUE),
  range(CampCreek_2$data$datetime, na.rm = TRUE)
))

ylim <- range(c(
  range(CampCreek_1$data[,'temperature'], na.rm = TRUE),
  range(CampCreek_2$data[,'temperature'], na.rm = TRUE)
))

# Now plot the temperature timeseries to look for overlaps 
# (use pch = 0 to speed up graphical rendering)
plot(CampCreek_1$data[,c('datetime', 'temperature')], 
     xlim = xlim, ylim = ylim, col = 'black', pch = '.',
     main = "Camp Creek")
points(CampCreek_2$data[,c('datetime', 'temperature')], 
       col = 'orange', pch = '.')
```

### Timeseries data 2 -- Non-overlapping timeseries

This next example shows non-overlapping timeseries

```{r timeseries_non_identical}
# MARIANNA/ST. FRANCSIS stations are separated by 56.7 m, row #5 from tooCloseTbl
#    ...
#    5   120   179     65.9

nwsID_1 <- meta$nwsID[120]
nwsID_2 <- meta$nwsID[179]

# Get timeseries data
Marianna <- fw13_createTimeseriesObject(nwsID_1, meta)
St_Francis <- fw13_createTimeseriesObject(nwsID_2, meta)

# Each timeseries object consists of 'meta' and 'data'.
# We are only interested in the 'data' part:

# ----- Plot them separately -----

layout(matrix(seq(2)))
plot(Marianna$data[,c('datetime', 'temperature')], 
     col = 'black', pch = '.', main = "Marianna")
plot(St_Francis$data[,c('datetime', 'temperature')], 
     col = 'orange', pch = '.', main = "St. Francis")
layout(1)

# ----- Plot them together -----

# Set up x- and y-limits to cover both timeseries
xlim <- range(c(
  range(Marianna$data$datetime, na.rm = TRUE),
  range(St_Francis$data$datetime, na.rm = TRUE)
))

ylim <- range(c(
  range(Marianna$data[,'temperature'], na.rm = TRUE),
  range(St_Francis$data[,'temperature'], na.rm = TRUE)
))

# Now plot the temperature timeseries to look for overlaps 
# (use pch = 0 to speed up graphical rendering)
plot(Marianna$data[,c('datetime', 'temperature')], 
     xlim = xlim, ylim = ylim, col = 'black', pch = '.')
points(St_Francis$data[,c('datetime', 'temperature')], 
       col = 'orange', pch = '.', main = "Marianna/St. Francis")
```


