---
title: "September 2020 Oregon Wildfires"
author: "Eli Grosman, Mazama Science"
date: "September 18, 2020"
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background
At the end of a hot and dry summer, the conditions were perfect for several large
wildfires to spread across western Oregon. The purpose of this document is to 
demonstrate the **RAWSmet** package's data gathering and modeling capabilities
by investigating the factors that allowed for the wildfires to spread. 

## Assembling data
The following code is executed to gather data from all of the WRCC stations 
in Oregon from the dates surrounding the start of the Holiday Farm fires:
```{r assembleData, message = FALSE}
library(RAWSmet)

setRawsDataDir("~/Data/RAWS")

oregonMeta <- wrcc_loadMeta(stateCode = "OR")

filePath <- file.path(getRawsDataDir(), "stationList_OR.rda")

if ( file.exists(filePath) ) {
  
  stationList_OR <- get(load(filePath))
  
} else {
  
  stationList_OR <- list()
  counter <- 1
  
  for ( wrccID in oregonMeta$wrccID ) {
    
    message(sprintf("Working on %s (%d/%d)", wrccID, counter, nrow(oregonMeta)))
    
    # Keep going even if a particular station doesn't have data
    res <- try({
      
      stationTimeseriesObject <- 
        wrcc_load(wrccID = wrccID, meta = oregonMeta, year = 2020) %>%
        raws_filterDate(
          startdate = MazamaCoreUtils::parseDatetime(20200831, timezone = "America/Los_Angeles"),
          enddate = MazamaCoreUtils::parseDatetime(20200914, timezone = "America/Los_Angeles"))
      
      stationList_OR[[wrccID]] <- stationTimeseriesObject
    }, silent = TRUE)
    
    counter <- counter + 1
      
  }
  
  save(stationList_OR, file = filePath)
    
}
```

## Locations with data
Across the state of Oregon there are 205 RAWS stations with data provided by the 
WRCC, however, only 134 of these stations have accessible data from the dates of 
the fires. The locations of these stations in addition to station metdata
can be seen on the following map: 
```{r oregonStations, message = FALSE}
wrccIDs <- names(stationList_OR)

meta <- oregonMeta %>% dplyr::filter(wrccID %in% wrccIDs)
RAWSmet::raws_leaflet(meta)
```

## Wind shift maps
These maps show the direction and speeds of winds and gusts at all of the available
stations in Oregon. It is clear that over the dates of the fires there are 
strong winds coming from the east. 
```{r windShift, message = FALSE, echo = FALSE, fig.align = "center", fig.width = 10, fig.height = 10}
library(maps)

par(mfrow = c(2, 2))

maps::map('state', 'oregon', fill = FALSE, col = 'gray80')
title("Sept. 6th at noon")

for ( station in stationList_OR ) {
  data <- 
    station$data %>%
    dplyr::filter(
      datetime == MazamaCoreUtils::parseDatetime(2020090612, timezone = "America/Los_Angeles")
    )
  knots <- data$maxGustSpeed * 1.944
  addWindBarbs(
    station$meta$longitude, 
    station$meta$latitude, 
    knots, 
    data$maxGustDirection,
    circleSize = 0.5,
    circleFill = 'black',
    lineCol = 1,
    extraBarbLength = -0.2,
    lwd = 1.2
  )
}


maps::map('state', 'oregon', fill = FALSE, col = 'gray80')
title("Sept. 6th at midnight")

for ( station in stationList_OR ) {
  data <- 
    station$data %>%
    dplyr::filter(
      datetime == MazamaCoreUtils::parseDatetime(2020090700, timezone = "America/Los_Angeles")
    )
  knots <- data$maxGustSpeed * 1.944
  addWindBarbs(
    station$meta$longitude, 
    station$meta$latitude, 
    knots, 
    data$maxGustDirection,
    circleSize = 0.5,
    circleFill = 'black',
    lineCol = 1,
    extraBarbLength = -0.2,
    lwd = 1.2
  )
}


maps::map('state', 'oregon', fill = FALSE, col = 'gray80')
title("Wind Gusts on Sept. 7th at noon")

for ( station in stationList_OR ) {
  data <- 
    station$data %>%
    dplyr::filter(
      datetime == MazamaCoreUtils::parseDatetime(2020090712, timezone = "America/Los_Angeles")
    )
  knots <- data$maxGustSpeed * 1.944
  addWindBarbs(
    station$meta$longitude, 
    station$meta$latitude, 
    knots, 
    data$maxGustDirection,
    circleSize = 0.5,
    circleFill = 'black',
    lineCol = 1,
    extraBarbLength = -0.2,
    lwd = 1.2
  )
}


maps::map('state', 'oregon', fill = FALSE, col = 'gray80')
title("Wind Gusts on Sept. 7th at midnight")

for ( station in stationList_OR ) {
  data <- 
    station$data %>%
    dplyr::filter(
      datetime == MazamaCoreUtils::parseDatetime(2020090800, timezone = "America/Los_Angeles")
    )
  knots <- data$maxGustSpeed * 1.944
  addWindBarbs(
    station$meta$longitude, 
    station$meta$latitude, 
    knots, 
    data$maxGustDirection,
    circleSize = 0.5,
    circleFill = 'black',
    lineCol = 1,
    extraBarbLength = -0.2,
    lwd = 1.2
  )
}
```

## Wind shift time series plots
Lets take a closer look at some stations in western Oregon. 
```{r windTimeseriesPlots, message = FALSE, echo = FALSE, fig.align = "center", fig.height = 10, fig.width = 10}

# Get data

orOBRR <- 
  wrcc_load(wrccID = "orOBRR", meta = oregonMeta, year = 2020) %>%
  raws_filterDate(20200904, 20200908)

orOGOO <- 
  wrcc_load(wrccID = "orOGOO", meta = oregonMeta, year = 2020) %>%
  raws_filterDate(20200904, 20200908)

orOBRRData <- orOBRR %>% raws_extractData()
orOGOOData <- orOGOO %>% raws_extractData()

# Create plots

orOBRRwindTimeseries <- windTimeseriesPlot(orOBRR)
orOGOOwindTimeseries <- windTimeseriesPlot(orOGOO)

orOBRRtimeseriesMultiplot <-
  timeseriesMultiplot(orOBRRData, pattern = c("humidity|temperature"), nrow = 2) + 
  ggplot2::ggtitle(sprintf("Temperature and Humidity in %s, %s", 
                           orOBRR$meta$wrccID, orOBRR$meta$siteName)) +
  ggplot2::xlab("Time") +
  ggplot2::ylab("Temperature (°C), Humidity (%)") +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))

orOGOOtimeseriesMultiplot <-
  timeseriesMultiplot(orOGOOData, pattern = c("humidity|temperature"), nrow = 2) +
  ggplot2::ggtitle(sprintf("Temperature and Humidity in %s, %s", 
                           orOGOO$meta$wrccID, orOGOO$meta$siteName)) +
  ggplot2::xlab("Time") +
  ggplot2::ylab("Temperature (°C), Humidity (%)") +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))

gridExtra::grid.arrange(orOBRRwindTimeseries, orOGOOwindTimeseries, orOBRRtimeseriesMultiplot, orOGOOtimeseriesMultiplot)
```
It is apparent that over labor day weekend, these stations experienced high winds 
from the east in addition to high temperatures and low humidity. 

## Extra

