---
title: "September 2020 Oregon Wildfires"
author: "Eli Grosman, Mazama Science"
date: "September 18, 2020"
output: rmarkdown::html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background
At the end of a hot and dry summer, the conditions were set for several large
wildfires to spread across western Oregon. The purpose of this document is to 
demonstrate the **RAWSmet** package's data gathering and modeling capabilities
by investigating the factors that allowed for the wildfires to spread. 

## Assembling data
First, we will use **RAWSmet**'s `wrcc_loadMeta()` and `wrcc_loadMultiple()` functions 
to gather recent data from all of the stations in Oregon. The following code 
is executed to gather the necessary data between August 31st and September 14th,
approximately a week before and after the start of the wildfires in Holiday Farm.
```{r assembleData, message = FALSE}
library(RAWSmet)

setRawsDataDir("~/Data/RAWS")

# Load WRCC station metadata for Oregon
oregonMeta <- wrcc_loadMeta(stateCode = "OR")

filePath <- file.path(getRawsDataDir(), "stationList_OR.rda")

# Load pre-generated data if it exists. If it does not exist, create it.
if ( file.exists(filePath) ) {
  
  stationList_OR <- get(load(filePath))
  
} else {
  
  # Create required data
  stationList_OR <- 
    wrcc_loadMultiple(wrccIDs = oregonMeta$wrccID, meta = oregonMeta, year = 2020, password = MY_PASSWORD) %>%
    rawsList_filterDate(
      startdate = MazamaCoreUtils::parseDatetime(20200831, timezone = "America/Los_Angeles"),
      enddate = MazamaCoreUtils::parseDatetime(20200914, timezone = "America/Los_Angeles"))
  
  # Save this data so it does not need to be generated again in the future
  save(stationList_OR, file = filePath)

}
```

## Locations with data
Across the state of Oregon there are 205 RAWS stations with data provided by the 
WRCC, however, only 134 of these stations have accessible data from the dates of 
the fires. The locations of these stations in addition to station metdata
can be seen on the following map: 
```{r oregonStations, message = FALSE}
wrccIDs <- names(stationList_OR)

meta <- oregonMeta %>% dplyr::filter(wrccID %in% wrccIDs)
RAWSmet::raws_leaflet(meta)
```

## Wind shift maps
With all of the necessary data assembled, we can now take a look at the wind patterns
around the dates of the wildfires. These maps, created with `maps::map()` and 
`RAWSMet::rawsList_addWindBarbs()`, show the direction and speeds of *wind gusts* at 
all of the available stations in Oregon. 

Through these maps it is apparent that on September 6th at noon there is no irregular pattern
in the wind that stands out. However, at midnight on the 6th, the eastern side of
Oregon experienced strong winds coming from the north. Over the night and throughout
the morning, these winds move and the western side of the Cascades
is seen to have strong winds coming from the east. 
```{r windShift, message = FALSE, echo = FALSE, fig.align = "center", fig.width = 10, fig.height = 10}
library(maps)

par(mfrow = c(2, 2))

maps::map('state', 'oregon', fill = FALSE, col = 'gray80')
title("Sept. 6th at noon")

rawsList_addWindBarbs(
  stationList_OR,
  observationTime = MazamaCoreUtils::parseDatetime(2020090612, timezone = "America/Los_Angeles"),
  skipMissing = TRUE,
  type = "gusts",
  circleSize = 0.5,
  circleFill = 'black',
  lineCol = 1,
  extraBarbLength = -0.2,
  lwd = 1.2
)

maps::map('state', 'oregon', fill = FALSE, col = 'gray80')
title("Sept. 6th at midnight")
  
rawsList_addWindBarbs(
  stationList_OR,
  observationTime = MazamaCoreUtils::parseDatetime(2020090700, timezone = "America/Los_Angeles"),
  skipMissing = TRUE,
  type = "gusts",
  circleSize = 0.5,
  circleFill = 'black',
  lineCol = 1,
  extraBarbLength = -0.2,
  lwd = 1.2
)

maps::map('state', 'oregon', fill = FALSE, col = 'gray80')
title("Sept. 7th at noon")

rawsList_addWindBarbs(
  stationList_OR,
  observationTime = MazamaCoreUtils::parseDatetime(2020090712, timezone = "America/Los_Angeles"),
  skipMissing = TRUE,
  type = "gusts",
  circleSize = 0.5,
  circleFill = 'black',
  lineCol = 1,
  extraBarbLength = -0.2,
  lwd = 1.2
)

maps::map('state', 'oregon', fill = FALSE, col = 'gray80')
title("Sept. 7th at midnight")

rawsList_addWindBarbs(
  stationList_OR,
  observationTime = MazamaCoreUtils::parseDatetime(2020090800, timezone = "America/Los_Angeles"),
  skipMissing = TRUE,
  type = "gusts",
  circleSize = 0.5,
  circleFill = 'black',
  lineCol = 1,
  extraBarbLength = -0.2,
  lwd = 1.2
)

```

## Wind shift time series plots
Lets take a closer look at some stations west of the cascades. With **RAWSmet**'s
`windTimeseriesPlot()` and `timeseriesMultiplot`, we can see the change in the
wind pattern and the change in temperature and humidity over this time period.
It is apparent that  over labor day weekend, the direction of the wind suddenly 
changed to point east and gust speeds increased considerably. Furthermore, 
these stations experienced an extreme drop in humidity and temperature increased
overall.
```{r windTimeseriesPlots, message = FALSE, echo = FALSE, fig.align = "center", fig.height = 10, fig.width = 10}

# Get data

orOBRR <- 
  stationList_OR[["orOBRR"]] %>%
  raws_filterDate(20200904, 20200908)

orOGOO <- 
  stationList_OR[["orOGOO"]] %>%
  raws_filterDate(20200904, 20200908)

orOBRRData <- orOBRR %>% raws_extractData()
orOGOOData <- orOGOO %>% raws_extractData()

# Create plots

orOBRRwindTimeseries <- windTimeseriesPlot(orOBRR)
orOGOOwindTimeseries <- windTimeseriesPlot(orOGOO)

orOBRRtimeseriesMultiplot <-
  timeseriesMultiplot(orOBRRData, pattern = c("humidity|temperature"), nrow = 2) + 
  ggplot2::ggtitle(sprintf("Temperature and Humidity in %s, %s", 
                           orOBRR$meta$wrccID, orOBRR$meta$siteName)) +
  ggplot2::xlab("Time") +
  ggplot2::ylab("Temperature (°C), Humidity (%)") +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))

orOGOOtimeseriesMultiplot <-
  timeseriesMultiplot(orOGOOData, pattern = c("humidity|temperature"), nrow = 2) +
  ggplot2::ggtitle(sprintf("Temperature and Humidity in %s, %s", 
                           orOGOO$meta$wrccID, orOGOO$meta$siteName)) +
  ggplot2::xlab("Time") +
  ggplot2::ylab("Temperature (°C), Humidity (%)") +
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5))

gridExtra::grid.arrange(orOBRRwindTimeseries, orOGOOwindTimeseries, orOBRRtimeseriesMultiplot, orOGOOtimeseriesMultiplot)
```

## Wind rose plots
Using **openair**'s `windRose()` plot, we can take a closer look at the wind 
patterns in western Oregon. The following plots compare the winds at Goodwin Peak 
in September 2020 to historical data for the winds in each September between 2005
and 2017. From the historical data it is clear that the winds in Goodwin Peak in 
September typically come from the north. However, in September 2020, an unmistakable
majority of the winds come from the east.
```{r windRoseAllTime, echo = FALSE, message = FALSE}
library(openair)

fw13Meta <- fw13_loadMeta()

# Get all stations with valid data on the west side of the cascades
meta <- fw13Meta %>%
  dplyr::filter(stateCode == "OR") %>%
  dplyr::filter(longitude < -122.33)

allData_westernOregon <- data.frame()

for ( i in 1:nrow(meta) ) {
  res <- try({
    stationTSObject <- fw13_load(nwsID = meta[i,]$nwsID, meta = fw13Meta)
    
    # Get data for each September
    stationData <- stationTSObject %>%
      raws_extractData(forOpenair = TRUE) %>%
      dplyr::filter(lubridate::month(datetime) == 9)
    
    allData_westernOregon <- rbind(allData_westernOregon, stationData)
  }, silent = TRUE)
}

# openair::windRose(allData_westernOregon, ws = "windSpeed", wd = "windDirection", main = "Wind direction and speed. Western oregon, all time.")
```

```{r windRosesOGOO, echo = FALSE, message = FALSE, fig.width = 5, fig.height = 5, fig.show = 'hold'}

par(mfrow = c(1, 2))

filePath <- file.path(getRawsDataDir(), "orOGOO_wrcc_0831-0919")

# Load pre-generated data if it exists. If it does not exist, create it.
if ( file.exists(filePath) ) {
  
  wrccData_orOGOO <- get(load(filePath))

} else {
  
  # Create data
  wrccData_orOGOO <- 
    wrcc_loadYear(wrccID = "orOGOO", year = 2020, password = MY_PASSWORD) %>%
    raws_filterDate(
      startdate = MazamaCoreUtils::parseDatetime(20200901, timezone = "America/Los_Angeles"),
      enddate = MazamaCoreUtils::parseDatetime(20200930, timezone = "America/Los_Angeles")
    ) %>%
    raws_extractData(forOpenair = TRUE)
  
  save(wrccData_orOGOO, file = filePath)
}


fw13Data_orOGOO <-
  fw13_load(nwsID = "352545") %>%
  raws_extractData(forOpenair = TRUE) %>%
  dplyr::filter(lubridate::month(datetime) == 9)

fw13Plot <- openair::windRose(mydata = fw13Data_orOGOO,
                              ws = "windSpeed",
                              wd = "windDirection",
                              main = "Goodwin peak, September 2005-2017.")

wrccPlot <- openair::windRose(mydata = wrccData_orOGOO,
                              ws = "windSpeed",
                              wd = "windDirection",
                              main = "Goodwin peak, September 2020.")
```
