---
title: "Raw Data Validation"
author: "Jonathan Callahan"
date: "3/12/2022"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

unitBreaks <- function(x) { return(ceiling(max(x, na.rm = TRUE) - min(x, na.rm = TRUE))) }
```

In ths document, we will examine raw data obtained from 
https://cefa.dri.edu/raws/fw13/ and do some basic exploratory statitics to 
better understand any data quality issues that might exist.

----

# Download and Parse NWSID 040203

Here, we use the **RAWSmet** package to download and parse data for a single
location. No cleanup or QC is performed. We are only bringing the data into 
memory in R so we can analyze it.

```{r download_parse, message = FALSE}
library(dplyr)
library(RAWSmet)
setRawsDataDir("~/Data/RAWS")


# Load data for NWSID 040203 -- "BLUE RIDGE (KNF)"
tbl <-
  cefa_downloadData(nwsID = "040203") %>%
  cefa_parseData()

# Show variables
names(tbl)
```

In order to plot data, we need a valid `POSIXct` time axis. We create one here
for a station in California reporting in local time.

```{r time_axis}
datetime <- 
  paste0(tbl$observationDate, tbl$observationTime) %>%
  MazamaCoreUtils::parseDatetime(timezone = "America/Los_Angeles")
```

# Dry-bulb Temeprature

## Time series plot

The first thing to do with any variable is simply look at it. We plot it here
with different values for partial opacity to better see overplotted values.

```{r temp_timeseries, fig.height = 10, results = "hold"}
layout(matrix(seq(2)))
plot(datetime, tbl$dryBulbTemp, pch = 15, cex = 0.4, col = adjustcolor("black", 0.1))
plot(datetime, tbl$dryBulbTemp, pch = 15, cex = 0.4, col = adjustcolor("black", 0.004))
layout(1)
```

In the upper plot, our eyes notice a few lone outliers
and a an outbreak of outliers in the winter of 2014.

In the lower plot, we can see a _'line'_ of overplotted values near 32 F.

## Distribution of values

Every atmospheric measurement has reasonable physical limits to its range. 
Temperature in the continental US has rarely, if ever exceeded -50 F to +130 F.

Now let use a histogram and box plot to better see the distribution of values.

```{r distribution_of_values, results = "hold"}
layout(matrix(seq(2)))
hist(tbl$dryBulbTemp, n = unitBreaks)
boxplot(tbl$dryBulbTemp, horizontal = TRUE)
layout(1)
```

So far, so good. We have a reasonably smooth distribution of values with a few
outliers and an expected spike in readings near the freezing point of water. Temperature
should be expected to hover briefly near this phase change point, as latent heat
associated with freezing or melting water is not measured by the dry-bulb temperature.

## Rate of change

Every atmospheric variable also has bounds on its rate of change. We can examine
that distribution a well.

```{R rate_of_change, results = "hold"}
layout(matrix(seq(2)))
x <- diff(tbl$dryBulbTemp)
hist(x, n = unitBreaks)
boxplot(x, horizontal = TRUE)
layout(1)
```

This distribution looks reasonable even if some of values (> 50F/hr) seem
questionable.

# Zero = missing value flag

The question was raised as to whether zero was used as a missing value flag in
for some variables. Lets find out how many times the temperature was exactly
zero:

```{r count_zeroes}
length(which(tbl$dryBulbTemp == 0))
```

For temperature, this does not seem to be an issue.

# Summary

Except for a few individual outliers and a couple of outlier outbreaks, the
values found for "Dry-bulb temperature" appear reasonable.


