---
title: "Working With openair"
author: "Eli Grosman, Mazama Science"
date: "August 18, 2020"
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

The **RAWSmet** package was developed to make downloading and 
working with weather data gathered by RAWS stations simpler. Using this package 
in conjunction with the **openair** package can make some beautiful and 
informative visualizations of RAWS data.

[**Openair**](https://davidcarslaw.github.io/openair/) is an R package designed 
specifically for modeling air quality data. The package provides many different 
tools for plotting wind and pollution roses, flexible time series plots, and more. 
Additionally, **openair** can easily group data and plot it by different periods 
such as by the hour, day, day of the week, season, and year. More information 
and documentation for **openair** may be found on its website: 
https://davidcarslaw.github.io/openair/.

The goal of this document is to introduce the use of **RAWSmet** and **openair** 
to create visualizations of weather data.. 

## Setting up RAWSmet

Follow these instructions to set up **RAWSmet** correctly. These instruction may 
also be found on the package’s website: https://mazamascience.github.io/RAWSmet/.

To follow along with the rest of this document you may also install **openair** 
by running the following in the RStudio console:

```
install.packages("openair")
```

The **RAWSmet** package is designed to be used with [R](https://cran.r-project.org) 
(>= 3.5) and [RStudio](https://www.rstudio.com) so make sure you have those 
installed first.

Users will want to install the **remotes** package to have access to the latest 
version of the package from GitHub.

The following packages should be installed by typing the following at the 
RStudio console:

```
# Note that vignettes require knitr and rmarkdown
install.packages('knitr')
install.packages('rmarkdown')
install.packages('MazamaSpatialUtils')
remotes::install_github('MazamaScience/MazamaLocationUtils')
remotes::install_github('MazamaScience/RAWSmet')
```
Any work with spatial data, *e.g.* assigning states, counties and timezones, 
will require installation of required spatial datasets. To get these datasets 
you should type the following at the RStudio console:

```
library(MazamaSpatialUtils)
dir.create('~/Data/Spatial', recursive = TRUE)
setSpatialDataDir('~/Data/Spatial')
installSpatialData()
```

Data generated with package functions can be be saved to and reloaded from a
dedicated directory much the same as the `spatialDataDir` used above. Set up 
a directory for RAWS data with:

```
library(RAWSmet)
dir.create('~/Data/RAWS', recursive = TRUE)
setRawsDataDir('~/Data/RAWS')
```

The `rawsDataDir` must be set up correctly to use its functionality. RAWS data
takes a long time to download so data may be saved to this directory so it does
not need to be downloaded again in the future.

## Preparing data for openair

Throughout this section, we will be using **RAWSmet**'s `fw13_load()` and 
`fw13_loadMeta()` functions. These functions will either load the specified 
data from `rawsDataDir` or download and save it to the `rawsDataDir` if it 
does not exist. You may also use `wrcc_loadYear()` and `wrcc_loadMeta()` for RAWS
data from the WRCC. 

Before using RAWS data with **openair**, we must ensure that the data is 
downloaded correctly and is in the correct format. 

First, create or load all of the FW13 station metadata using `fw13_loadMeta()`.

```{r fw13_createTimeseriesObject, message = FALSE}
library(RAWSmet)
library(MazamaSpatialUtils)

setSpatialDataDir("~/Data/Spatial")
setRawsDataDir("~/Data/RAWS")

meta <- fw13_loadMeta()
head(meta)
```

The `raws_leaflet()` function provides an interactive map to help find the ID 
associated with a particular location. Just click on a dot to find the associated
ID:

```{r raws_leaflet}
raws_leaflet(meta)
```

We will choose the station in Enumclaw, Washington (451702) and create or load a
"timeseries object". This timeseries object contains two dataframes, one of 
station metadata and another of cleaned weather data.

```{r view_ts_object}
# nwsID 451702 is Enumclaw, WA
enumclawTSObj <- fw13_load(nwsID = 451702, meta = meta)

names(enumclawTSObj)
# View station metadata
enumclawTSObj$meta
# View sample of raw data
head(enumclawTSObj$data)
```

Note that this RAWS timeseries object contains *all* of the data gathered by the 
specified station in its lifetime. We can filter the data to look at periods of 
interest using `raws_filterDate()`. This function can understand any date that is
understood by `lubridate::ymd()`.

Also note that the data stored in these RAWS timeseries objects are in UTC. 

```{r filter_dates, message = FALSE}

# 20050101 will be parsed as Jan. 1st, 2005
# 20060101 will be parsed as Jan 1st. 2006

# Get all observations between these dates
enumclawTSObj_2005 <- 
  raws_filterDate(enumclawTSObj, 
                  startdate = 20050101,
                  enddate = 20060101,
                  timezone = "America/Los_Angeles")

range(enumclawTSObj_2005$data$datetime)

# 20050801 will be parsed as Aug. 1st, 2005
# 20050901 will be parsed as Sep. 1st, 2005

# Get all observations between these dates
enumclawTSObj_200508 <- 
  raws_filterDate(enumclawTSObj,
                  startdate = 20050801,
                  enddate = 20060901,
                  timezone = "America/Los_Angeles")

range(enumclawTSObj_200508$data$datetime)
```

**Openair** requires that dates and times of observations are stored in a column 
called `date`. However, **RAWSmet** names this column `datetime`. Use `raws_extractData()`
with `forOpenair = TRUE` to extract the `data` dataframe from a RAWS timeseries 
object with a new column called `date` containing the same values as `datetime`.

```{r rename_datetime, message = FALSE}


enumclawData_2005 <-raws_extractData(enumclawTSObj_2005, forOpenair = TRUE)
enumclawData_200508 <-raws_extractData(enumclawTSObj_2005, forOpenair = TRUE)

```

All of the functions used above may be used separately as demonstrated but they
can also be strung neatly together by using the pipe `%>%` operator. The same data
may be generated like so:
```{r generate_with_pipe, message = FALSE}

meta <- fw13_loadMeta()

enumclawData_2005 <- 
  fw13_load(nwsID = 451702, meta = meta) %>%
  raws_filterDate(20050101, 20060101, timezone = "America/Los_Angeles") %>%
  raws_extractData(forOpenair = TRUE)

enumclawData_200508 <- 
  fw13_load(nwsID = 451702, meta = meta) %>%
  raws_filterDate(20050801, 20050901, timezone = "America/Los_Angeles") %>%
  raws_extractData(forOpenair = TRUE)

# We will also extract a dataframe of ALL of the station's data
enumclawData_ALL <-
  fw13_load(nwsID = 451702, meta = meta) %>%
  raws_extractData(forOpenair = TRUE)
```

After renaming the datetime column and filtering by dates of interest, we are 
now ready to create visualizations using **openair.** 

## Using openair

The RAWS timeseries data contains measurements for temperature, humidity, wind speed, 
wind direction, max gust speed, max gust direction, precipitation, and solar 
radiation. We can utilize various **openair** plots to gain insight from each 
of these parameters.

### Wind rose plots

Let us first create some wind rose plots. **Openair**'s `windRose()` function 
requires 3 arguments: the data to create the plot for, and the names of the wind 
speed and wind direction columns. By default, `windRose()` looks for columns 
named `ws` and `wd` for wind speed and direction respectively so it is important 
to specify these names when calling the function. (_Remember that **openair** 
requires that dates and times of observations be stored in a column named `date`._)

```{r windRose, message = FALSE}
library(openair)

openair::windRose(
  enumclawData_200508, 
  ws = "windSpeed", 
  wd = "windDirection",
  main = "Wind speed and direction in Enumclaw, August 2005", 
  key.footer = "(mph)"
)
```

`windRose()` can also group data and plot it by different periods. Lets look at 
the wind speed and directions by season in Enumclaw:

```{r windRose_seasons}
openair::windRose(
  enumclawData_2005, 
  ws = "windSpeed", 
  wd = "windDirection",
  main = "Wind speed and direction in Enumclaw, 2005", 
  type = "season", 
  key.footer = "(mph)"
)
```

### Time-series plotting

**Openair** can also be used to create time-series plots and trends. Lets first 
take a look at the function `timePlot()`. This function requires 2 arguments: 
the data to create the plot for, and `pollutant`, the name of the column to plot 
with respect to time.  (_Again, `timePlot()` requires that dates and times of 
observations are stored in a column named `date`._)

```{r temp_timePlot}
openair::timePlot(
  enumclawData_200508, 
  pollutant = "temperature",
  avg.time = "hour",
  main = "Temperature in Enumclaw, August 2005", 
  key = FALSE, 
  xlab = "time",
  ylab = "temperature (°F)"
)
```

`timePlot()` can also plot multiple columns so they can be compared against each 
other. Lets compare temperature and humidity in Enumclaw in August 2005:

```{r temp_humid_timePlot}
openair::timePlot(
  enumclawData_200508, 
  pollutant = c("temperature", "humidity"),
  avg.time = "hour",
  main = "Temperature and Humidity in Enumclaw, August 2005", 
  key = TRUE,
  name.pol = c("temperature (°F)", "humidity (%)"), 
  ylab = ""
)
```

Plotting trends in data is also very easy using **openair.** The `smoothTrend()` 
function plots monthly averages against the trend in the variable of interest. 
Lets look at the trend of solar radiation in Enumclaw in 2005:

```{r smoothTrend_months}
openair::smoothTrend(
  enumclawData_2005, 
  pollutant = "solarRadiation",
  avg.time = "month",
  main = "Solar Radiation trend in Enumclaw, 2005", 
  statistic = "mean",
  xlab = "time",
  ylab = expression('solar radiation (W/m'^2*')')
)
```

Instead of comparing monthly averages to the trend of the data, **openair** can 
also compare different averages.

```{r smoothTrend_years}
openair::smoothTrend(
  enumclawData_ALL,  
  pollutant = "solarRadiation",
  main = "Solar Radiation trend in Enumclaw", 
  statistic = "mean",
  xlab = "time", 
  ylab = expression('solar radiation (W/m'^2*')'), 
  avg.time = "year"
)
```
