---
title: "Working With openair"
author: "Eli Grosman, Mazama Science"
date: "August 18, 2020"
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

The **RAWSmet** package was developed to make downloading and 
working with weather data gathered by RAWS stations simpler. Using this package 
in conjunction with the **openair** package can make some beautiful and 
informative visualizations of RAWS data.

[**Openair**](https://davidcarslaw.github.io/openair/) is an R package designed 
specifically for modeling air quality data. The package provides many different 
tools for plotting wind and pollution roses, flexible time series plots, and more. 
Additionally, **openair** can easily group data and plot it by different periods 
such as by the hour, day, day of the week, season, and year. More information 
and documentation for **openair** may be found on its website: 
https://davidcarslaw.github.io/openair/.

The goal of this document is to introduce the use of **RAWSmet** and **openair** 
to create visualizations of weather data.. 

## Setting up RAWSmet

Follow these instructions to set up **RAWSmet** correctly. These instruction may 
also be found on the package’s website: https://mazamascience.github.io/RAWSmet/.

To follow along with the rest of this document you may also install **openair** 
by running the following in the RStudio console:

```
install.packages("openair")
```

The **RAWSmet** package is designed to be used with [R](https://cran.r-project.org) 
(>= 3.5) and [RStudio](https://www.rstudio.com) so make sure you have those 
installed first.

Users will want to install the **remotes** package to have access to the latest 
version of the package from GitHub.

The following packages should be installed by typing the following at the 
RStudio console:

```
# Note that vignettes require knitr and rmarkdown
install.packages('knitr')
install.packages('rmarkdown')
install.packages('MazamaSpatialUtils')
remotes::install_github('MazamaScience/MazamaLocationUtils')
remotes::install_github('MazamaScience/RAWSmet')
```
Any work with spatial data, *e.g.* assigning states, counties and timezones, 
will require installation of required spatial datasets. To get these datasets 
you should type the following at the RStudio console:

```
library(MazamaSpatialUtils)
dir.create('~/Data/Spatial', recursive = TRUE)
setSpatialDataDir('~/Data/Spatial')
installSpatialData()
```

<!-- Data generated with package functions can be be saved to and reloaded from a -->
<!-- dedicated directory much the same as the `spatialDataDir` used above. Set up -->
<!-- a directory for RAWS data with: -->

<!-- ``` -->
<!-- library(RAWSmet) -->
<!-- dir.create('~/Data/RAWS', recursive = TRUE) -->
<!-- setRawsDataDir('~/Data/RAWS') -->
<!-- ``` -->

<!-- The `rawsDataDir` must be set up correctly to use its functionality. RAWS data  -->
<!-- takes a long time to download so data may be saved to this directory so it does  -->
<!-- not need to be downloaded again in the future. -->

## Preparing data for openair

Before using RAWS data with **openair**, we must ensure that the data is 
downloaded correctly and is in the correct format. 

First, create a RAWS timeseries object using **RAWSmet**’s 
`fw13_createTimeseriesObject()` function (or `wrcc_createTimeseriesObject()` for 
current RAWS data from the WRCC). 

```{r fw13_createTimeseriesObject, message = FALSE}
library(RAWSmet)
library(MazamaSpatialUtils)

setSpatialDataDir("~/Data/Spatial")

meta <- fw13_createMetadata()
head(meta)
```

The `raws_leaflet()` function provides an interactive map to help find the ID 
associated with a particular location. Just click on a dot to find the associated
ID:

```{r raws_leaflet}
raws_leaflet(meta)
```

We will choose the station in Enumclaw, Washington (451702) and create a 
"timeseries object". This timeseries object contains two dataframes, one of 
station metadata and another of cleaned weather data.

```{r view_ts_object}
# nwsID 451702 is Enumclaw, WA
enumclawTSObj <- fw13_createTimeseriesObject(nwsID = 451702, meta = meta)

names(enumclawTSObj)
# View station metadata
enumclawTSObj$meta
# View sample of raw data
head(enumclawTSObj$data)
```

**Openair** requires that dates and times of observations are stored in a column 
called `date`. However, **RAWSmet** names this column `datetime`. Create the 
correct date column by running:

```{r rename_datetime, message = FALSE}
library(dplyr)

# Create a new column by copying the data from datetime
enumclawTSObj$data$date <- enumclawTSObj$data$datetime

# Another solution is to rename the column using dplyr:
# enumclawTSObj$data <- dplyr::rename(enumclawTSObj$data, date = datetime)
```

Note that this RAWS timeseries object contains *all* of the data gathered by the 
specified station in its lifetime. We can filter the data to look at periods of 
interest. It is also important to note that the datetimes stored in these RAWS 
timeseries objects are of class `POSIXct` and are in the UTC timezone. 

The `MazamaCoreUtils::parseDatetime()` function provides a robust datetime
parser that enforces use of an explicit timezone. We can use this function to
provide some start and end dates and then use normal **dplyr** filtering to subset
the dataframe:

```{r filter_dates, message = FALSE}
library(MazamaCoreUtils)
# 20050101 will be parsed as Jan. 1st, 2005
start2005 <- MazamaCoreUtils::parseDatetime(20050101, timezone = "UTC")
# 20060101 will be parsed as Jan 1st. 2006
end2005 <- MazamaCoreUtils::parseDatetime(20060101, timezone = "UTC")

# 20050801 will be parsed as Aug. 1st, 2005
startAug2005 <- MazamaCoreUtils::parseDatetime(20050801, timezone = "UTC")
# 20050901 will be parsed as Sep. 1st, 2005
endAug2005 <- MazamaCoreUtils::parseDatetime(20050901, timezone = "UTC")

# Get all observations between these dates
enumclawData_2005 <- dplyr::filter(enumclawTSObj$data, date >= start2005 & date < end2005)
range(enumclawData_2005$date)

enumclawData_200508 <- dplyr::filter(enumclawTSObj$data, date >= startAug2005 & date < endAug2005)
range(enumclawData_200508$datetime)
```

After renaming the datetime column and filtering by dates of interest, we are 
now ready to create visualizations using **openair.** 

## Using openair

The RAWS timeseries data contains measurements for temperature, humidity, wind speed, 
wind direction, max gust speed, max gust direction, precipitation, and solar 
radiation. We can utilize various **openair** plots to gain insight from each 
of these parameters.

### Wind rose plots

Let us first create some wind rose plots. **Openair**'s `windRose()` function 
requires 3 arguments: the data to create the plot for, and the names of the wind 
speed and wind direction columns. By default, `windRose()` looks for columns 
named `ws` and `wd` for wind speed and direction respectively so it is important 
to specify these names when calling the function. (_Remember that **openair** 
requires that dates and times of observations be stored in a column named `date`._)

```{r windRose, message = FALSE}
library(openair)

openair::windRose(
  enumclawData_200508, 
  ws = "windSpeed", 
  wd = "windDirection",
  main = "Wind speed and direction in Enumclaw, August 2005", 
  key.footer = "(mph)"
)
```

`windRose()` can also group data and plot it by different periods. Lets look at 
the wind speed and directions by season in Enumclaw:

```{r windRose_seasons}
openair::windRose(
  enumclawData_2005, 
  ws = "windSpeed", 
  wd = "windDirection",
  main = "Wind speed and direction in Enumclaw, 2005", 
  type = "season", 
  key.footer = "(mph)"
)
```

### Time-series plotting

**Openair** can also be used to create time-series plots and trends. Lets first 
take a look at the function `timePlot()`. This function requires 2 arguments: 
the data to create the plot for, and `pollutant`, the name of the column to plot 
with respect to time.  (_Again, `timePlot()` requires that dates and times of 
observations are stored in a column named `date`._)

```{r temp_timePlot}
openair::timePlot(
  enumclawData_200508, 
  pollutant = "temperature",
  avg.time = "hour",
  main = "Temperature in Enumclaw, August 2005", 
  key = FALSE, 
  xlab = "time",
  ylab = "temperature (°F)"
)
```

`timePlot()` can also plot multiple columns so they can be compared against each 
other. Lets compare temperature and humidity in Enumclaw in August 2005:

```{r temp_humid_timePlot}
openair::timePlot(
  enumclawData_200508, 
  pollutant = c("temperature", "humidity"),
  avg.time = "hour",
  main = "Temperature and Humidity in Enumclaw, August 2005", 
  key = TRUE,
  name.pol = c("temperature (°F)", "humidity (%)"), 
  ylab = ""
)
```

Plotting trends in data is also very easy using **openair.** The `smoothTrend()` 
function plots monthly averages against the trend in the variable of interest. 
Lets look at the trend of solar radiation in Enumclaw in 2005:

```{r smoothTrend_months}
openair::smoothTrend(
  enumclawData_2005, 
  pollutant = "solarRadiation",
  avg.time = "month",
  main = "Solar Radiation trend in Enumclaw, 2005", 
  statistic = "mean",
  xlab = "time",
  ylab = expression('solar radiation (W/m'^2*')')
)
```

Instead of comparing monthly averages to the trend of the data, **openair** can 
also compare different averages.

```{r smoothTrend_years}
openair::smoothTrend(
  enumclawTSObj$data, 
  pollutant = "solarRadiation",
  main = "Solar Radiation trend in Enumclaw", 
  statistic = "mean",
  xlab = "time", 
  ylab = expression('solar radiation (W/m'^2*')'), 
  avg.time = "year"
)
```
